Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-26 article.tex
minor changes

Given this assumption, we then propose to use as a measure of the
strength of the connection between two neurons $i$ and $j$, the
partial correlation coefficient $p_{i,j}$ between their corresponding
random variables $X_i$ and $X_j$, defined by \cite{}:
\begin{equation}
p_{i,j} =
-\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}, \label{eq:inverse}
\end{equation}
where $\Sigma^{-1}$, known as the precision or concentration matrix, is the inverse of the covariance matrix $\Sigma$ of $X$. %%  Partial
%% correlation can be interpreted in several ways. The partial
%% correlation coefficient $p_{i,j}$ can be shown to measure the
%% correlation between $X_i$ and $X_j$ when they are corrected for the
%% (linear) effect of all other variables in $X$ \cite{}.
Assuming that the distribution $P_X$ is a multivariate Gaussian
distribution ${\cal N}(\mu,\Sigma)$, it can be shown that $p_{i,j}$ is
zero if and only if $X_i$ and $X_j$ are independent given all other
variables in $X$, ie. $X_i \perp X_j|X^{-i,j}$ where $X^{-i,j}= X
\setminus\{X_i,X_j\}$. Partial correlation thus measures conditional
dependencies between variables. As an association measures between
neurons, it should thus naturally only detect direct associations
between neurons and filter out spurious indirect effects. The interest
of partial correlation as an association measure has already been
shown for the inference of gene regulatory networks \cite{xxx}.
%% Given a set of multivariate Gaussian
%% random variables $X$, the (undirected) graph that connects all pairs
%% of variables $X_i$ and $X_j$ such that $p_{i,j}\neq 0$ is called a
%% Gaussian graphical models (GGMs). Because it measures conditional
%% dependencies, partial correlation should naturally only detect direct
%% associations between neurons. The superiority of partial correlation
%% over standard correlation as an association measure have already been
%% shown in the context of the inference of gene regulatory networks from
%% gene expression data \cite{}. We will compare both measures in Section
%% \ref{sec:results}.
Note that the partial correlation statistic is symmetric
(i.e. $p_{i,j}=p_{j,i}$). Therefore, our approach cannot identify the
direction of the interactions between neurons. We will see in
Section~\ref{sec:results} why this only affects slightly its
performances, with respect to the metric used in the Connectomics
challenge.

@@

The top of Table \ref{tab:comparison} reports AUROC and AUPRC for all four networks
using in each case partial correlation with different filtering
functions. Except for the last two rows that uses PCA, the exact
inverse of the covariance matrix was used in each case. These results
clearly show the importance of the filters. AUROC increases in average from
0.77 to 0.93. PCA does not really affect AUROC scores but it improves
significantly AUPRC scores. Taking the average over various parameter
setting gives an improvement of 10\% \textcolor{red}{AS:AUPRC ok but AUROC?} in both AUROC and AUPRC. The last
row shows the final performance of the method specifically tuned for
the challenge (see the Appendix \ref{app:optimized} for all details). Although this tuning
was decisive to obtain the best performance at the challenge, it does
not significantly improve neither AUROC nor AUPRC. Note that tuning
was mainly done on the \textit{normal-1} and \textit{normal-4}
datasets, which explains why these two networks are slightly better
retrieved than the others.

@@

In the bottom of Table \ref{tab:comparison}, we provide as a comparison the
performance of three other methods: standard (Pearson) correlation (PC),
generalized transfer entropy (GTE), and GENIE3. ROC and PR curves on the
\textit{normal-2} network are shown for all methods in Figure
\ref{fig:}. Pearson correlation measures the unconditional linear
(in)dependence between variables and it should thus not be able to filter out
indirect interactions between neurons. GTE \citep{stetter2012model} was
proposed as a baseline for the challenge. This method builds on Transfer
Entropy to measure the association between two neurons. Unlike our approach, it
can predict the direction of the edges. GENIE3 \citep{huynhthu2010inferring} is
a gene regulatory network inference method that was the best performer of the
DREAM5 challenge \cite{dream5paper}. When transposed to neural networks, this
method uses as a confidence score for the edge going from neuron $i$ to neuron
$j$ the importance score of variable $X_i$ in a Random Forests model trying to
predict $X_j$ from all variables in $X\setminus X_j$. To reduce computing times
of this method, we had to reduce each tree in the Random Forests model to a
maximum depth of 3. This constraint potentially severly affects the performance
of this method with respect to the use of fully grown trees. We nevertheless
provide these results for comparison purpose. PC and GENIE3 were both applied
on the same pre-filtered time series ($g \circ f_1$). For GENIE3, we
built 10000 trees per neuron and we used default settings for all other
parameters (except for the maximal tree depth). For GTE, we reproduced the
exact same setting (conditioning level and pre-processing) that was used for
the challenge.

@@

Partial correlation and average partial correlation clearly outperforms all
other methods. The improvement is more important in terms of AUPRC than in
terms of AUROC. As expected, Pearson correlation performs very poorly in terms
of AUPRC. GTE and GENIE3 works much better but these two methods are
nevertheless clearly below partial correlation. Among these two methods, GTE is
slightly better in terms of AUROC, while GENIE3 is significantly better in
terms of AUPRC. Given that we had to limit this latter method for computational
reasons, these results are very promising and a comparison with the full GENIE3
approach is certainly part of our future works.

>>>>

Given this assumption, we then propose to use as a measure of the
strength of the connection between two neurons $i$ and $j$, the
partial correlation coefficient $p_{i,j}$ between their corresponding
random variables $X_i$ and $X_j$, defined by:
\begin{equation}
p_{i,j} =
-\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}, \label{eq:inverse}
\end{equation}
where $\Sigma^{-1}$, known as the precision or concentration matrix, is the inverse of the covariance matrix $\Sigma$ of $X$. %%  Partial
%% correlation can be interpreted in several ways. The partial
%% correlation coefficient $p_{i,j}$ can be shown to measure the
%% correlation between $X_i$ and $X_j$ when they are corrected for the
%% (linear) effect of all other variables in $X$ \cite{}.
Assuming that the distribution $P_X$ is a multivariate Gaussian
distribution ${\cal N}(\mu,\Sigma)$, it can be shown that $p_{i,j}$ is
zero if and only if $X_i$ and $X_j$ are independent given all other
variables in $X$, ie. $X_i \perp X_j|X^{-i,j}$ where $X^{-i,j}= X
\setminus\{X_i,X_j\}$. Partial correlation thus measures conditional
dependencies between variables. As an association measures between
neurons, it should thus naturally only detect direct associations
between neurons and filter out spurious indirect effects. The interest
of partial correlation as an association measure has already been
shown for the inference of gene regulatory networks
\cite{de2004discovery,Schafer:2005}.
%% Given a set of multivariate Gaussian
%% random variables $X$, the (undirected) graph that connects all pairs
%% of variables $X_i$ and $X_j$ such that $p_{i,j}\neq 0$ is called a
%% Gaussian graphical models (GGMs). Because it measures conditional
%% dependencies, partial correlation should naturally only detect direct
%% associations between neurons. The superiority of partial correlation
%% over standard correlation as an association measure have already been
%% shown in the context of the inference of gene regulatory networks from
%% gene expression data \cite{}. We will compare both measures in Section
%% \ref{sec:results}.
Note that the partial correlation statistic is symmetric
(i.e. $p_{i,j}=p_{j,i}$). Therefore, our approach cannot identify the
direction of the interactions between neurons. We will see in
Section~\ref{sec:results} why this only affects slightly its
performances, with respect to the metric used in the Connectomics
challenge.

@@

The top of Table \ref{tab:comparison} reports AUROC and AUPRC for all
four networks using in each case partial correlation with different
filtering functions. Except for the last two rows that uses PCA, the
exact inverse of the covariance matrix was used in each case. These
results clearly show the importance of the filters. AUROC increases in
average from 0.77 to 0.93. PCA does not really affect AUROC scores but
it improves significantly AUPRC scores. Taking the average over
various parameter setting gives an improvement of 10\% in AUPRC but
only a minor change in AUROC. The last row shows the final performance
of the method specifically tuned for the challenge (see the Appendix
\ref{app:optimized} for all details). Although this tuning was
decisive to obtain the best performance at the challenge, it does not
significantly improve neither AUROC nor AUPRC.

@@

In the bottom of Table \ref{tab:comparison}, we provide as a comparison the
performance of three other methods: standard (Pearson) correlation (PC),
generalized transfer entropy (GTE), and GENIE3. ROC and PR curves on the
\textit{normal-2} network are shown for all methods in Figure
\ref{fig:curves}. Pearson correlation measures the unconditional linear
(in)dependence between variables and it should thus not be able to filter out
indirect interactions between neurons. GTE \citep{stetter2012model} was
proposed as a baseline for the challenge. This method builds on Transfer
Entropy to measure the association between two neurons. Unlike our approach, it
can predict the direction of the edges. GENIE3 \citep{huynhthu2010inferring} is
a gene regulatory network inference method that was the best performer of the
DREAM5 challenge \citep{marbach2012}. When transposed to neural networks, this
method uses as a confidence score for the edge going from neuron $i$ to neuron
$j$ the importance score of variable $X_i$ in a Random Forests model trying to
predict $X_j$ from all variables in $X\setminus X_j$. To reduce the
computational cost of this method, we had to limit each tree in the Random
Forests model to a maximum depth of 3. This constraint potentially severly
affects the performance of this method with respect to the use of fully grown
trees. We nevertheless provide these results for comparison purpose. PC and
GENIE3 were both applied on the same pre-filtered time series ($g \circ
f_1$). For GENIE3, we built 10000 trees per neuron and we used default settings
for all other parameters (except for the maximal tree depth). For GTE, we
reproduced the exact same setting (conditioning level and pre-processing) that
was used for the challenge.

@@

Partial correlation and average partial correlation clearly outperform all
other methods. The improvement is more important in terms of AUPRC than in
terms of AUROC. As expected, Pearson correlation performs very poorly in terms
of AUPRC. GTE and GENIE3 works much better but these two methods are
nevertheless clearly below partial correlation. Among these two methods, GTE is
slightly better in terms of AUROC, while GENIE3 is significantly better in
terms of AUPRC. Given that we had to limit this latter method for computational
reasons, these results are very promising and a comparison with the full GENIE3
approach is certainly part of our future works.
